---
title: "Net Gain of Bikeshare Data"
author: "Sean DeSantis, Lucas Adams, Ryan Myles-Ferguson"
date: "2023-11-15"
output: pdf_document
---
# Loading packages
```{r}
library(dplyr)
library(tidyverse)
library(here)
library(janitor)
library(sf)
library(tmap)
library(tidycensus)
library(lubridate)
library(knitr)
```

# Loading data
```{r}
df1 <- read_csv(here("data_raw", "202309-capitalbikeshare-tripdata.csv"))
```

# Creating new dataframe that takes out any rows with missing information
```{r}
df1b = na.omit(df1)
```

# Subtracting end time from start time and finding the values where the start time exceeds end time and filtering them out of the data set

```{r}
dfb = df1b %>%
  mutate(ride_time = ended_at-started_at) %>%
  arrange(ride_time) %>%
  filter(ride_time>=0)
  
```

# Combining timestamps into one column
```{r}
df2b = pivot_longer(dfb, cols = c("started_at", "ended_at"), values_to = "time")
```

# New Variable for Start and End
```{r}
df3b = df2b %>%
  mutate(time_type = case_when(
  time %in% df1b$started_at ~ "start",       
  time %in% df1b$ended_at ~ "end",       
  ))
```

# Counts how many times a bike started at a station
```{r}
df_count_start = df3b %>%
  group_by(start_station_id, start_station_name, .drop = FALSE) %>%
  count(time_type = "start", name = "bikes_start") 
print(df_count_start)
```

# Counts how many times a bike ended at a station
```{r}
df_count_end = df3b %>%
  group_by(end_station_id, end_station_name, .drop = FALSE) %>%
 count(time_type = "end", name = "bikes_end")

print(df_count_end)
```
# Combines counts into one dataframe and computes the net gain of bikes at a station
```{r}
df4b=bind_cols(df_count_start,df_count_end) %>%
  mutate(net_gain = bikes_end - bikes_start) %>%
arrange(net_gain) 

print(df4b)
```

```{r}
df5b = df4b %>%
  mutate(station_id = coalesce(start_station_id, end_station_id)) %>%
  select(-start_station_id, -end_station_id) 
```

```{r}
df5b %>%
  ggplot(aes(x = net_gain, y = station_id)) + geom_point()
```


```{r}
neigh=st_read(here("data_raw", "DC_Health_Planning_Neighborhoods.geojson")) %>%
  clean_names()
class(neigh)
```

```{r}
df_c=read_csv(here("data_raw", "DC_COVID-19_Total_Positive_Cases_by_Neighborhood.csv")) %>%
clean_names()
```

```{r}
census_api_key("3e36eb0352f836410ded716bd5fc36a281dd8188")

```