---
title: "Playing With net_gain"
name: "Sean DeSantis"
output: html_notebook
---

```{r}
library(knitr)
library(tidyverse)
library(janitor)
library(here)
library(sf)
library(tmap)
```

```{r}
capital_bikeshare <- read_csv(here("data_raw", "202309-capitalbikeshare-tripdata.csv"))
```
# Creating new dataframe that takes out any rows with missing information
```{r}
capital_bikeshare_na = na.omit(capital_bikeshare)
```

# Subtracting end time from start time and finding the values where the start time exceeds end time and filtering them out of the data set

```{r}
capital_bikeshare_rt = capital_bikeshare_na %>%
  mutate(ride_time = ended_at-started_at) %>%
  arrange(ride_time) %>%
  filter(ride_time>=0)
  
```

# Combining timestamps into one column
```{r}
capital_bikeshare_rt2 = pivot_longer(capital_bikeshare_rt, cols = c("started_at", "ended_at"), values_to = "time")
```

# New Variable for Start and End
```{r}
capital_bikeshare_rt3 = capital_bikeshare_rt2 %>%
  mutate(time_type = case_when(
  time %in% capital_bikeshare_na$started_at ~ "start",       
  time %in% capital_bikeshare_na$ended_at ~ "end",       
  ))
```

# Counts how many times a bike started at a station
```{r}
bikeshare_rt_count_start = capital_bikeshare_rt3 %>%
  group_by(start_station_id, start_station_name, .drop = FALSE) %>%
  count(time_type = "start", name = "bikes_start") %>%
print(bikeshare_rt_count_start)
```

# Counts how many times a bike ended at a station
```{r}
bikeshare_rt_count_end = capital_bikeshare_rt3 %>%
  group_by(end_station_id, end_station_name, .drop = FALSE) %>%
 count(time_type = "end", name = "bikes_end")

print(bikeshare_rt_count_end)
```

# Combines counts into one dataframe and computes the net gain of bikes at a station
```{r}
bikeshare_rt_net_gain=bind_cols(bikeshare_rt_count_start,bikeshare_rt_count_end) %>%
  mutate(net_gain = bikes_end - bikes_start) %>%
arrange(net_gain) 

print(bikeshare_rt_net_gain)
```

```{r}
net_gain2 = select(.bikshare_rt_count_end, "bikes_end") %>%

net_gain = bikeshare_rt_count_end$bikes_end - bikeshare_rt_count_start$bikes_start
```

```{r}
bikeshare_rt_net_gain2 %>%
  group_by(bikshare_rt_count_end$end_station_name, bikeshare_rt_count_start$start_station_name)
  mutate(net_gain = bikeshare_rt_count_end$bikes_end - bikeshare_rt_count_start$bikes_start)
```


