---
title: "In Class Exercise: 18-10-2023"
name: "Lucas, Ryan, Sean"
date: "18-10-2023"
output: html_notebook
---

# Loading Packages
```{r}
library(tidyverse)
library(dplyr)
library(sf)
library(janitor)
library(here)
```

# Loading Data
```{r}
bikeshare_sept23 <- read.csv(here("data_raw", "202309-capitalbikeshare-tripdata.csv"))
```

# Slicing off first 10000 points to examine
```{r}
bikeshare_sept2302 = bikeshare_sept23  |>
  slice(1:100000) 
```

# Plotting Points on standard grid
```{r}
  ggplot(bikeshare_sept2302, mapping = aes(x = start_lng, y = start_lat))+
  geom_point()
```

# Preparing points for map plotting
```{r}
points = st_as_sf(bikeshare_sept2302, coords = c("start_lng", "start_lat"), crs = 4326)
plot(st_geometry(points), pch = 16, col = "navy")
```

# Plotting all points on the D.C. map
```{r}
library(OpenStreetMap)

upperLeft = c(39.12, -77.38)

lowerRight = c(38.77, -76.86)

base_map  = openmap(upperLeft, lowerRight, type="osm")

plot(base_map)

points = st_transform(points, osm())

plot(st_geometry(points), pch=16, col="navy", cex=0.5, add=T)
```
# Separating timestamps to access hour
```{r}
bikeshare_sept2302 <- bikeshare_sept2302 %>%
  separate(started_at, into = c("start_date", "start_time"), sep = 10, remove = FALSE)
bikeshare_sept2302 <- bikeshare_sept2302 %>%
  separate(ended_at, into = c("end_date", "end_time"), sep = 10, remove = FALSE)
bikeshare_sept2302 <- bikeshare_sept2302 %>%
  separate(start_time, into = c("start_hour", "start_inhour"), sep = 3, remove = FALSE)
bikeshare_sept2302 <- bikeshare_sept2302 %>%
  separate(end_time, into = c("end_hour", "end_inhour"), sep = 3, remove = FALSE)
```

# Mutating a new variable to turn string segment into a numeric value
```{r}
bikeshare_sept2302 <- bikeshare_sept2302 %>%
  mutate(begin_hr = case_when(
    start_hour == " 01" ~ 01,
    start_hour == " 02" ~ 02,
    start_hour == " 03" ~ 03,
    start_hour == " 04" ~ 04,
    start_hour == " 05" ~ 05,
    start_hour == " 06" ~ 06,
    start_hour == " 07" ~ 07,
    start_hour == " 08" ~ 08,
    start_hour == " 09" ~ 09,
    start_hour == " 10" ~ 10,
    start_hour == " 11" ~ 11,
    start_hour == " 12" ~ 12,
    start_hour == " 13" ~ 13,
    start_hour == " 14" ~ 14,
    start_hour == " 15" ~ 15,
    start_hour == " 16" ~ 16,
    start_hour == " 17" ~ 17,
    start_hour == " 18" ~ 18,
    start_hour == " 19" ~ 19,
    start_hour == " 20" ~ 20,
    start_hour == " 21" ~ 21,
    start_hour == " 22" ~ 22,
    start_hour == " 23" ~ 23,
    start_hour == " 00" ~ 00
  ))
```

# Grouping hours into morning and afternoon
```{r}
bikeshare_sept2302 <- bikeshare_sept2302 %>%
  mutate(tod = case_when(
  begin_hr %in% c(0,1,2,3,4,5) ~ 1,
  begin_hr %in% c(6,7,8,9,10,11) ~ 2,
  begin_hr %in% c(12,13,14,15,16,17) ~ 3,
  begin_hr %in% c(18,19,20,21,22,23) ~ 4
    ))
```

# Preparing points to be plotted again
```{r}
points = st_as_sf(bikeshare_sept2302, coords = c("start_lng", "start_lat"), crs = 4326)
plot(st_geometry(points), pch = 16, col = "navy")
```
# Plotting points again on D.C. map, this time distinguishing them by color, where red is morning and black is afternoon.
```{r}
palette = c("red", "black", "blue", "green")

names(palette) = unique(points$tod)

upperLeft = c(39.13, -77.38)

lowerRight = c(38.77, -76.84)

base_map  = openmap(upperLeft, lowerRight, type="osm")

plot(base_map)

points = st_transform(points, osm())

plot(st_geometry(points), pch=16, col=palette[points$tod], cex=0.5, add=T)
```
