---
title: "Intro to Data Science Final Project"
author: "Lucas Adams, Ryan Miles-Ferguson, Sean DeSantis"
output: html_notebook
---

```{r}
library(knitr)
library(tidyverse)
library(janitor)
library(here)
library(sf)
library(tmap)
```

```{r}
df1 <- read_csv(here("data_raw", "202309-capitalbikeshare-tripdata.csv"))
```
# Creating new dataframe that takes out any rows with missing information
```{r}
df1b = na.omit(df1)
```

# Subtracting end time from start time and finding the values where the start time exceeds end time and filtering them out of the data set

```{r}
dfb = df1b %>%
  mutate(ride_time = ended_at-started_at) %>%
  arrange(ride_time) %>%
  filter(ride_time>=0)
  
```

# Combining timestamps into one column
```{r}
df2b = pivot_longer(dfb, cols = c("started_at", "ended_at"), values_to = "time")
```

# New Variable for Start and End
```{r}
df3b = df2b %>%
  mutate(time_type = case_when(
  time %in% df1b$started_at ~ "start",       
  time %in% df1b$ended_at ~ "end",       
  ))
```

# Counts how many times a bike started at a station
```{r}
df_count_start = df3b %>%
  group_by(start_station_id, start_station_name, .drop = FALSE) %>%
  count(time_type = "start", name = "bikes_start") 
print(df_count_start)
```

# Counts how many times a bike ended at a station
```{r}
df_count_end = df3b %>%
  group_by(end_station_id, end_station_name, .drop = FALSE) %>%
 count(time_type = "end", name = "bikes_end")

print(df_count_end)
```
# Combines counts into one dataframe and computes the net gain of bikes at a station
```{r}
df4b=bind_cols(df_count_start,df_count_end) %>%
  mutate(net_gain = bikes_end - bikes_start) %>%
arrange(net_gain) 

print(df4b)
```

# We'll pause with this for a minute, as we need to be able to map everything first. First, we are going to load in the data for the location of each metro station and bikeshare dock.
```{r}
bikeshare=st_read(here("data_raw","Capital_Bikeshare_Locations.geojson"))
```

```{r}
metro = st_read(here("data_raw", "Metro_Stations_Regional.geojson"))
```

# Now, we'll plot the data just to get a quick sense of what everything looks like.
```{r}
plot(metro)
```

```{r}
plot(bikeshare)
```

# Next, we'll merge the location data to be able to effectively plot it.
```{r}
bikeshare_j=bind_rows(bikeshare,metro)
```

# When the data is merged, the station type (metro/bikeshare) is lost, so we'll use the OBJECTID to be able to identify each.
```{r}
bikeshare_metro = bikeshare_j %>% mutate(loc_type = case_when(
  OBJECTID <= 1000 ~ "Metro",
  OBJECTID > 1000 ~ "Bikeshare"
))
```

# Now we will plot the location of each metro station and each bikeshare dock using the tmap package.
```{r}
tmap_mode("view")
tm_shape(bikeshare_metro) + tm_dots("loc_type") + tm_text("NUM_BIKES_AVAILABLE",size = 0.5)
```

# Now that we have a good idea of where the stations are, we'll join the ridership data for each bikeshare station with the location data for the bikeshare data only.
```{r}
#df4b = df4b %>% rename(NAME = start_station_name)
bikeshare_squared = left_join(bikeshare, df4b, by = "NAME")
```

```{r}
bikeshare_squared_b = bikeshare_squared  %>%
  select(-STATION_STATUS, -GIS_ID) %>% drop_na()
#bikeshare_squared_b = na.omit(bikeshare_squared2)
```

```{r}
tmap_mode("view")
tm_shape(bikeshare_squared_b) + tm_dots("NAME", legend.show = FALSE) + tm_text("net_gain",size = 0.5)
```

```{r}
bikeshare_1000 = bikeshare_squared_b %>% filter(net_gain >= abs(50))
tmap_mode("view")
tm_shape(bikeshare_1000) + tm_dots("NAME", legend.show = FALSE) + tm_text("net_gain",size = 0.5)
```

```{r}
bikeshare_metro_1000 = bind_rows(bikeshare_1000, metro)
bikeshare_metro_1000 = bikeshare_metro_1000 %>% mutate(loc_type = case_when(
  OBJECTID <= 1000 ~ "Metro",
  OBJECTID > 1000 ~ "Bikeshare"
))
tmap_mode("view")
tm_shape(bikeshare_metro_1000) + tm_dots("loc_type")

```

How active was the Rider on a day-to-day basis?
```{r}
library(tidyr)
library(tidyverse)
# Create a table with just the datetime stamp
dfb_time = dfb |> select(started_at)
dfb_time2 = separate(dfb_time, col=started_at, into=c("day","time"), sep= " ")
dfb_time3 = separate(dfb_time2, col=day, into=c("year", "month", "day"), sep="-")
dfb_time4 = dfb_time3 |> count(day)

ggplot(dfb_time4, aes(x=day, y=n)) + geom_point() + ggtitle("Number of Riders per")
        
```
Base on the graph shown above, we can see that on 09/23. We experience the lowest Ridership where under 4,000 active Riders. This is the outliner on the graph, because our second lowest is 8,000 and on the day after this. Which is 09/24. This is interesting because how did the Ridership drop so much in such little time. What factors can lead to this result? Could this be weather, holidays that happen this is day?